preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution

### 2021-05-29 BMC_Genome_Biology_submission/grubert_manuscript_figures_tables/manuscript/manuscript_v1.Rmd
Start the new version

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of the 3-dimensional (3D) structure of the human genome such as chromatin loops [@rao20143d], Topologically Associating Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At the kilobase scale, chromatin loops connect gene promoters with distal enhancers, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. At the megabase scale, TADs represent regions on the linear genome that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation [@phillips2013architectural]. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Thus, identifying the precise location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify genomic coordinates of TADs (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised]. However, a key limitation of them is that they depend on Hi-C data *resolution*. Resolution refers to the size of genomic regions (bins) used to segment the linear genome [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which leads to increased uncertainty in the location of boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. While these low-resolution Hi-C datasets can be used to probe higher-order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TAD and loop boundaries [@fudenberg2016formation]. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth [@schmitt2016genome], obtaining high-resolution Hi-C data and boundaries remains difficult.

>>>>

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of the 3-dimensional (3D) structure of the human genome such as chromatin loops [@Lupianez:2015aa; @Franke:2016aa; @rao20143d; @Phanstiel:2017uq], Topologically Associating Domains (TADs) [@nora2012spatial; @dixon2012topological; @Hou:2012aa; @sexton2012three], and A/B compartments [@lieberman2009comprehensive; @rao20143d], reviewed in [@Beagan:2020wz]. At the kilobase scale, chromatin loops (corner-dot structures on Hi-C chromatin interaction maps) connect gene promoters with distal enhancers, promoting regulation of gene expression. At the megabase scale, TADs represent regions on the linear genome that are highly self-interacting. Perhaps the most prominent feature of TADs is that they are demarcated by boundaries, suggesting that most of the mammalian genome is folded into non-overlapping chromatin interaction domains [@nora2012spatial; @dixon2012topological; @Hou:2012aa; @sexton2012three]. Boundaries were found to be enriched in architectural factors such as CTCF (in convergent orientation [@rao20143d; @Tang:2015aa]), RAD21, SMC3 (members of the cohesin complex), YY1, and ZNF143 [@Dixon:2012aa; @Rao:2014aa; @Phillips-Cremins2013:aa; @Corces:2016aa; @Heidari:2014wr]. <!--Emerging evidence has linked TADs and loops to regulation of gene expression, cellular dynamics, differentiation [@phillips2013architectural; @Beagan:2020wz]. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa].--> Disruption of boundaries have been reported to promote human cancers [@hnisz2016activation; @taberlay2016three; @Flavahan2016], neurological disorders [@Sun:2018aa], and pathologies of limb development [@Franke:2016aa; @lupianez2016breaking]. Thus, identifying the precise location of boundaries remains a top priority to fully understand the functionality of the human genome.

TAD and loop boundaries are not mutually exclusive. Rao et al. demonstrated that boundaries of TADs, referred to as "contact domains", are enriched in chromatin loops [@rao20143d]. Numerous observations demonstrated the presence of hierarchically nested smaller chromatin domains within TADs [@phillips2013architectural; @rao20143d], also demarcated by shared boundaries [@Norton:2018aa; @Cresswell:2020ab]. Boundaries were shown to form by the seminal 'loop extrusion' mechanism of the molecular motors (condensin and cohesin) tracking along the DNA sequence, ‘extruding’ the intervening DNA, and pausing at the convergent CTCF motifs [@Alipour:2012aa; @Fudenberg:2016aa; @Goloborodko:2016vz]. The loop extrusion mechanism is now a leading hypothesis for the formation of distinct classes of chromatin loops and contact domains [@Beagan:2020wz]. In our work, we are focusing on TAD and loop boundaries created by the loop extrusion mechanism, and refer to the associated chromatin structures as "contact domains".

The low-resolution of Hi-C data is a primary limitation for precise boundary detection. Resolution refers to the size of genomic regions (bins) used to segment the linear genome [@li2019comparison]. Because increasing resolution of Hi-C data requires a quadratic increase in the total sequencing depth [@schmitt2016genome] and the associated costs, most available Hi-C datasets have relatively low resolution, ranging from 5 kb to 100 kb.<!--While these low-resolution Hi-C datasets can be used to probe higher-order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TAD and loop boundaries [@fudenberg2016formation].--> Lower resolution corresponding to larger bin sizes leads to increased uncertainty in boundary location.

In contrast to low-resolution Hi-C data, functional/regulatory genome annotations (histone modifications, DNAse I hypersensitive sites, DNA methylation, and transcription factor binding sites (TFBSs)) assembled by the ENCODE and other consortia have been profiled at relatively high resolution (10-300bp). The association of domain boundaries with CTCF and other architectural proteins [@rao20143d; @Tang:2015aa; @Phillips:2009aa] holds promise that these high-resolution marks may inform the more precise location of domain boundaries. Indeed, methods like Lollipop

The ENCODE, NIH Roadmap Epigenomics, FANTOM5, BLUEPRINT, and other members of the International Human Epigenome Consortium (IHEC) [@Stunnenberg:2016aa] have been actively cataloging functional/regulatory genome annotation datasets, such as cell type- and tissue-specific histone modifications, DNAse I hypersensitive sites, DNA methylation, and transcription factor binding sites (TFBSs).

Numerous studies demonstrated the association of domain boundaries with genomic annotations [@Dixon:2015aa; @Whalen:2016aa; @Dekker:2016aa; @Bonev:2016aa; @Wang:2016ab].

Several methods have been proposed to identify genomic coordinates of TADs (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised]. However, a key limitation of them is that they depend on Hi-C data *resolution*.
